---
title: "Calibration Curve"
output: 
    html_document:
        toc: true
        toc_float: true
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


```{r}
pkgs = c("dplyr", "reshape2", "tibble", "Metabase", "ggplot2", "readxl", "plotly",
         "latex2exp", "DT")
for(pkg in pkgs){
    suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}
```

```{r}
load("../../data/hdl.rda")
curve = glc$curve$curve_params
curve_data = glc$curve$curve_data
```

```{r}
plot_calibration_curve = function(protein){
    df = data.frame(
        resp = curve_data$resp[, protein],
        conc = curve_data$conc[, protein]
    )
    
    (ggplot(df, aes(x = resp, y = conc)) +
        stat_smooth(method = "lm", color = "steelblue") +
        geom_point(shape = 21, color = "white", fill = "black", size = 2) +
        annotate(
            "text",
            x = min(df$resp) + (max(df$resp - min(df$resp))) * 0.1,
            y = max(df$conc) - (max(df$conc - min(df$conc))) * 0.05,
            label = paste0("R^2 = ", round(curve[protein, "r2"], 5))
        ) +
        theme_bw()
        ) %>%
        ggplotly
}
plot_hist = function(feature){
    df = data.frame(
        resp = as.numeric(glc$peptide$conc_table[feature,])
    )
    (ggplot(df) +
        geom_histogram(aes(x = resp), color = "white") +
        labs(title = feature) +
        theme_bw()) %>%
        ggplotly
}
```


## 1. Information

This dataset has `r length(unique(glc$peptide$feature_data$Protein))` different proteins. `r sum(sapply(unique(glc$peptide$feature_data$Protein), function(prot){any(glc$glycoforms$feature_data$Protein == prot )}))` proteins have at least one glycoforms detected, while `r length(unique(glc$peptide$feature_data$Protein)) - sum(sapply(unique(glc$peptide$feature_data$Protein), function(prot){any(glc$glycoforms$feature_data$Protein == prot )}))` do not have any glycoforms.

```{r}
data.frame(
    Protein = unique(glc$peptide$feature_data$Protein)
) %>%
    mutate(
        `Number of Peptides` = sapply(Protein, function(prot){
            sum(glc$peptide$feature_data$Protein == prot)
        }),
        `Number of Glycoforms` = sapply(Protein, function(prot){
            sum(glc$glycoforms$feature_data$Protein == prot)
        })
    ) %>%
    arrange(desc(`Number of Glycoforms`)) %>%
    datatable()
```

Five Proteins have calibration curves. They are Clus, ApoC1, ApoA1, ApoD, and ApoE. The calculated calibration curve parameters are below:

```{r}
datatable(curve)
```


## 2. Calibration Curves

### 2.1. Clus

This calibration curve has a decent $R^2$ value. 

Calculated calibration parameters:

+ slop: `r curve["Clus", "slope"]`
+ intercept: `r curve["Clus", "intercept"]`

```{r}
plot_calibration_curve("Clus")
```

"The dataset only has one CLUS peptide, `QuantPep-CLUS_ASSIIDELFQDR Results`, which ranges from `r min(glc$peptide$conc_table["QuantPep-CLUS_ASSIIDELFQDR Results",])` to `r max(glc$peptide$conc_table["QuantPep-CLUS_ASSIIDELFQDR Results",])`, that is perfectly in the range of the calibration curve.

```{r}
plot_hist("QuantPep-CLUS_ASSIIDELFQDR Results")
```

### 2.2. ApoC1

This curve has a great $R^2$, with the calculated parameters:

+ slop: `r curve["ApoC1", "slope"]`
+ intercept: `r curve["ApoC1", "intercept"]`

```{r}
plot_calibration_curve("ApoC1")
```

There is also only one ApoC1, `pep-APOC1-QSELSAK Results`, ranges from `r min(glc$peptide$conc_table["pep-APOC1-QSELSAK Results",], na.rm = TRUE)` to `r max(glc$peptide$conc_table["pep-APOC1-QSELSAK Results",], na.rm = TRUE)`. There seems to be one person with an extreme high value.

```{r}
plot_hist("pep-APOC1-QSELSAK Results")
```

### 2.3. ApoA1

The ApoA1 curve does not seem to be very ideal. The calculated parameters are blow:

+ slop: `r curve["ApoA1", "slope"]`
+ intercept: `r curve["ApoA1", "intercept"]`


```{r}
plot_calibration_curve("ApoA1")
```

There are two ApoA1 peptides:

+ `pep-APOA1_LAEYHAK Results`, ranges from `r min(glc$peptide$conc_table["pep-APOA1_LAEYHAK Results",], na.rm = TRUE)` to `r max(glc$peptide$conc_table["pep-APOA1_LAEYHAK Results",], na.rm = TRUE)`
+ `QuantPep-APOA1_DLATVYVDVLK Results`, ranges from `r min(glc$peptide$conc_table["QuantPep-APOA1_DLATVYVDVLK Results",], na.rm = TRUE)` to `r max(glc$peptide$conc_table["QuantPep-APOA1_DLATVYVDVLK Results",], na.rm = TRUE)`

Both the peptides are in a range that is higher than the highest point of the calibration curve.

```{r}
plot_hist("pep-APOA1_LAEYHAK Results")
```

```{r}
plot_hist("QuantPep-APOA1_DLATVYVDVLK Results")
```

### 2.4. ApoD

This calibration curve is perfect.

+ slop: `r curve["ApoD", "slope"]`
+ intercept: `r curve["ApoD", "intercept"]`

```{r}
plot_calibration_curve("ApoD")
```

The ApoD peptide, `QuantPep-APOD_VLNQELR Results`, ranges from `r min(glc$peptide$conc_table["QuantPep-APOD_VLNQELR Results",], na.rm = TRUE)` to `r max(glc$peptide$conc_table["QuantPep-APOD_VLNQELR Results",], na.rm = TRUE)`

```{r}
plot_hist("QuantPep-APOD_VLNQELR Results")
```


### 2.5. ApoE

The ApoE calibration curve is also perfect. The parameters are:

+ slop: `r curve["ApoE", "slope"]`
+ intercept: `r curve["ApoE", "intercept"]`


```{r}
plot_calibration_curve("ApoE")
```

Three ApoE peptides are found from the dataset:

+ `pep-APOE_LGPLVEQGR_z3 Results`， ranges from `r min(glc$peptide$conc_table["pep-APOE_LGPLVEQGR_z3 Results",], na.rm = TRUE)` to `r max(glc$peptide$conc_table["pep-APOE_LGPLVEQGR_z3 Results",], na.rm = TRUE)`
+ `pep-APOE_LGPLVEQGR_z4 Results`， ranges from `r min(glc$peptide$conc_table["pep-APOE_LGPLVEQGR_z4 Results",], na.rm = TRUE)` to `r max(glc$peptide$conc_table["pep-APOE_LGPLVEQGR_z4 Results",], na.rm = TRUE)`
+ `APOE_AATVGSLAGQPLQER_z2 Results`, ranges from `r min(glc$peptide$conc_table["APOE_AATVGSLAGQPLQER_z2 Results",], na.rm = TRUE)` to `r max(glc$peptide$conc_table["APOE_AATVGSLAGQPLQER_z2 Results",], na.rm = TRUE)`

```{r}
plot_hist("pep-APOE_LGPLVEQGR_z3 Results")
```

```{r}
plot_hist("pep-APOE_LGPLVEQGR_z4 Results")
```

```{r}
plot_hist("APOE_AATVGSLAGQPLQER_z2 Results")
```

## 3. Missing Values

### Unglycosylated Peptides

The graph below shows the number of zeros in each unglycosylated peptide. 

```{r}
(table(apply(glc$peptide$conc_table, 1, function(row) sum(row == 0))) %>%
    as.data.frame %>%
    ggplot() +
    geom_col(aes(x = Var1, y = Freq)) +
    labs(x = "Number of Zeros", title = "Peptides") +
    theme_bw()) %>%
    ggplotly
```

The peptide with 21 zero is `r featureNames(glc$peptide)[apply(glc$peptide$conc_table, 1, function(row) sum(row == 0))==21]`. Fortunately the other two APOM peptdies do not have any zero

```{r}
subset_features(
    glc$peptide,
    grepl("APOM", featureNames(glc$peptide))
) %>%
    conc_table %>% apply(1, function(row) sum(row == 0))
```

The peptide with 13 zero is `r featureNames(glc$peptide)[apply(glc$peptide$conc_table, 1, function(row) sum(row == 0))==13]`. And again, the other APOC3 peptdies do not have any zero.

```{r}
subset_features(
    glc$peptide,
    grepl("APOC3", featureNames(glc$peptide))
) %>%
    conc_table %>% apply(1, function(row) sum(row == 0))
```

The peptide with 6 zero is `r featureNames(glc$peptide)[apply(glc$peptide$conc_table, 1, function(row) sum(row == 0))==6]`. And the other LCAT peptdies do not have any zero.

```{r}
subset_features(
    glc$peptide,
    grepl("LCAT", featureNames(glc$peptide))
) %>%
    conc_table %>% apply(1, function(row) sum(row == 0))
```

The peptide with 3 zero is `r featureNames(glc$peptide)[apply(glc$peptide$conc_table, 1, function(row) sum(row == 0))==3]`. 
```{r}
subset_features(
    glc$peptide,
    grepl("APOA", featureNames(glc$peptide))
) %>%
    conc_table %>% apply(1, function(row) sum(row == 0))
```

The two peptdies with 2 zerois are:

```{r}
featureNames(glc$peptide)[apply(glc$peptide$conc_table, 1, function(row) sum(row == 0))==1]
```

Both ApoA5 and APOL1 have only one unglycosylated peptide detected, and neither of them have any glycoforms. 

### Glycoforms

```{r}
(table(apply(glc$glycoforms$conc_table, 1, function(row) sum(row == 0))) %>%
    as.data.frame %>%
    ggplot() +
    geom_col(aes(x = Var1, y = Freq)) +
    labs(x = "Number of Zeros", title = "Glycoforms") +
    theme_bw()) %>%
    ggplotly
```

1. Peptides with more than 2 zeros will be removed.

2. Glycans with 6 or more zeros will be removed.

3. For the rest glycans, zeros will be filled with the lowest value in all samples devided by 2.

4. Glycans will be normalized to its unglycosylated peptide.

## 4. Peptides

```{r}
glc$peptide$feature_data %>%
    rownames_to_column("feature_id") %>%
    mutate(
        num_zero = apply(glc$peptide$conc_table, 1, function(row) sum(row == 0)),
        mean_response = rowMeans(glc$peptide$conc_table)
    ) %>%
    # mutate(
    #     mean_response = format(mean_response, digits = 3)
    # ) %>%
    datatable(options = list(order = list(2, "asc"))) %>%
    formatSignif(6, 3)
```


## 5. Summary

1. Many proteins have multiple peptides reported. Which peptide should we used to report the "abundance" of the particular protein? Should we use the one with the highest response (intensity)?

2. Which peptide should we use to calculate the concentration with the calibration curves?

3. The ApoA1 curve does not seem to be ideal 
